Black-swan events in animal populations

### 2016-11-24 text/ms_pnas.tex
Tiny edits

\begin{figure*}[htb]
\centering
\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
\caption{Illustration of population dynamic models that allow for heavy
tails. (A, B) Probability density for the Student-$t$
distribution with scale parameter of $1$ and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (C to E)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels C and D represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (F to H)
Estimates of \(\nu\) from models fit to the times series in panels C to
E. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind panels F to H
illustrates the approximate region of heavy tails.}
\label{fig:1}
\end{figure*}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.3\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (A to D)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel A are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

@@

One alternative approach would involve fitting
the generalized extreme value (GEV) distribution,
which represents the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
(e.g.\ sedimentation rates or wind speeds \cite{gaines1993, denny2009}),
the GEV requires maxima per time block.
Therefore, the GEV requires longer or higher frequency time series than typically available
for wild animal population abundance.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

@@

To generate five-year forecasts of abundance, we combined the posterior
parameter samples from the skewed-t Gompertz models with stochastically
generated process noise. We compared these forecasts to those generated by a
standard Gompertz model with normally distributed process noise fit to the same
data. We calculated the ratio of the lower 99\% quantile credible interval
between the two model projections for all populations in which Pr(\(\nu < 10\))
\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
posterior, we increased the number of posterior samples in Stan. We ran four
chains of 20,000 iterations and discarded the first 10,000 as warmup for a
total of 40,000 samples. We then generated five sets of future stochastic
process deviations for each for each sample
for a total of 200,000 forecast samples per population.

@@

\acknow{We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto,
  two anonymous reviewers and
members of the Earth to Ocean Group for helpful discussions and comments. We
are grateful to the contributors and maintainers of the Global Population
Dynamics Database and to Compute Canada's WestGrid high-performance computing
resources. Silhouette images were obtained from \url{phylopic.org} under Creative
Commons licenses; sources are listed in the Supporting Information. Funding was
provided by a Simon Fraser University Graduate Fellowship and David H. Smith
Conservation Research Fellowship (S.C.A.), the Natural Sciences and Engineering
Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program
(N.K.D.).
}

>>>>

\begin{figure*}[htb]
\centering
\includegraphics[width=0.63\textwidth]{../analysis/t-nu-eg2.pdf}
\caption{Illustration of population dynamic models that allow for heavy
tails. (A, B) Probability density for the Student-$t$
distribution with scale parameter of $1$ and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (C to E)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels C and D represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (F to H)
Estimates of \(\nu\) from models fit to the times series in panels C to
E. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind panels F to H
illustrates the approximate region of heavy tails.}
\label{fig:1}
\end{figure*}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.28\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (A to D)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel A are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

@@

One alternative approach would involve fitting
the generalized extreme value (GEV) distribution,
which represents the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for environmental variables
(e.g.\ sedimentation rates or wind speeds \cite{gaines1993, denny2009}),
the GEV requires maxima per time block.
Therefore, the GEV requires longer or higher frequency time series than typically available
for wild animal population abundance.
Furthermore, to our knowledge, the GEV cannot
be easily integrated into population dynamics models.

@@

To generate five-year forecasts of abundance, we combined the posterior
parameter samples from the skewed-t Gompertz models with stochastically
generated process noise. We compared these forecasts to those generated by a
standard Gompertz model with normally distributed process noise fit to the same
data. We calculated the ratio of the lower 99\% quantile credible interval
between the two model projections for all populations in which Pr(\(\nu < 10\))
\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
posterior, we increased the number of posterior samples in Stan. We ran four
chains of 20,000 iterations and discarded the first 10,000 as warmup.
We then generated five sets of future stochastic
process deviations for each for each sample
for a total of 200,000 forecast samples per population.

@@

\acknow{We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto,
  two anonymous reviewers, and
members of the Earth to Ocean Group for helpful discussions and comments. We
are grateful to the contributors and maintainers of the Global Population
Dynamics Database and to Compute Canada's WestGrid high-performance computing
resources. Silhouette images were obtained from \url{phylopic.org} under Creative
Commons licenses; sources are listed in the Supporting Information. Funding was
provided by a Simon Fraser University Graduate Fellowship and David H. Smith
Conservation Research Fellowship (S.C.A.), the Natural Sciences and Engineering
Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program
(N.K.D.).
}
